noam_data_exploration_v1

### 2024-07-19 noam_data_exploration_v3.Rmd
Added diversity stability at the quadrat level. looks great!

richness_mod <- glmmTMB(
  Richness ~ Island_side*Habitat + (1 | Island_side/Site/Transect/Location) + ar1(Year_factor + 0 | Location),
  family = poisson,
  data = alpha_diversity_quad,
  control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000))
) # overspecified but going with it for now
summary(richness_mod) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod)
plot(allEffects(richness_mod))

# probably the closest to what others have done:
richness_mod_2 <- glmmTMB(
  Richness ~ Year_factor*Habitat + (1 | Site/Transect/Location) + ar1(Year_factor + 0 | Location),
  family = poisson,
  data = alpha_diversity_quad,
  control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000))
) # overspecified but going with it for now
summary(richness_mod_2) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod_2)
plot(allEffects(richness_mod_2))
emmeans(richness_mod_2, pairwise ~ Habitat | Year_factor) # if we want to compare habitats year to year

@@

rate_change_untransformed_distances <- merge(rate_change_untransformed, rate_change_meta, by = "Location")

@@

rate_change_transformed_distances <- merge(rate_change_transformed, rate_change_meta, by = "Location")

@@

```{r diversity stability quadrat level setup}
data$Stratum <- as.factor(paste(data$Site, "-", data$Habitat, "-", data$Transect))
data$Location <- as.factor(data$Location)

quadrat_portfolio_columns <- c("Location", "Portfolio_effect") # names of columns for dataframe
quadrat_level_portfolio <- 
  data.frame(matrix(nrow = length(unique(data$Location)), 
                    ncol = length(quadrat_portfolio_columns))
             )
colnames(quadrat_level_portfolio) <- quadrat_portfolio_columns # change column 
quadrat_level_portfolio$Location <- unique(data$Location) # populate

for(i in 1:length(unique(unique(data$Location))) ){ # for each  quadrat
  # create a new dataframe for current quadrat
  newData <- subset(data, Location == levels(data$Location)[i])
  current_matrix <- newData[ , -which(names(newData) %in% meta_cols)] # remove metadata columns
  if(sum(current_matrix) == 0 | # if this was a 0 algae quadrat
     # or if there was only 1 column after removing all irrelevant algae
     is.null(ncol(current_matrix[, which(colSums(current_matrix) != 0)]))){ 
    quadrat_level_portfolio[i, 2] <- NA # set portfolio effects to 0
  }else{ # otherwise
    current_matrix <- current_matrix[, which(colSums(current_matrix) != 0)] # remove irrelevant algae

@@

sum(is.na(quadrat_level_portfolio$Portfolio_effect)) # 4 quadrats causing trouble
problem_children <- droplevels(quadrat_level_portfolio[is.na(quadrat_level_portfolio$Portfolio_effect),]$Location)
hist(quadrat_level_portfolio$Portfolio_effect) # some crazy outliers due to low richness

@@

>>>>

# richness_mod <- glmmTMB(
#   Richness ~ Island_side*Habitat + (1 | Island_side/Site/Transect/Location) + ar1(Year_factor + 0 | Location),
#   family = poisson,
#   data = alpha_diversity_quad,
#   control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000))
# ) # overspecified but going with it for now
# summary(richness_mod) # looks like strong autocorrelation, which makes sense to me
# Anova(richness_mod)
# plot(allEffects(richness_mod))

# probably the closest to what others have done:
richness_mod_2 <- glmmTMB(
  Richness ~ Year_factor*Habitat*Site + (1 | Site/Transect/Location) + ar1(Year_factor + 0 | Location),
  family = poisson,
  data = alpha_diversity_quad,
  control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000))
) # overspecified but going with it for now

summary(richness_mod_2) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod_2)
plot(allEffects(richness_mod_2))
# emmeans(richness_mod_2, pairwise ~ Habitat | Year_factor) # if we want to compare habitats year to year

@@

rate_change_untransformed_distances <- merge(rate_change_untransformed, turnover_meta, by = "Location")

@@

rate_change_transformed_distances <- merge(rate_change_transformed, turnover_meta, by = "Location")

@@

```{r diversity stability quadrat level setup, fig.height = 20, fig.width=20}
data_sub$Stratum <- as.factor(paste(data_sub$Site, "-", data_sub$Habitat, "-", data_sub$Transect))
data_sub$Location <- as.factor(data_sub$Location)

quadrat_portfolio_columns <- c("Location", "Portfolio_effect") # names of columns for dataframe
quadrat_level_portfolio <- 
  data.frame(matrix(nrow = length(unique(data_sub$Location)), 
                    ncol = length(quadrat_portfolio_columns))
             )
colnames(quadrat_level_portfolio) <- quadrat_portfolio_columns # change column 
quadrat_level_portfolio$Location <- unique(data_sub$Location) # populate

for(i in 1:length(unique(unique(data_sub$Location))) ){ # for each  quadrat
  # create a new dataframe for current quadrat
  newData <- subset(data, Location == levels(data_sub$Location)[i])
  current_matrix <- newData[ , -which(names(newData) %in% meta_cols)] # remove metadata columns
  if(sum(current_matrix) == 0 | # if this was a 0 algae quadrat
     # or if there was only 1 column after removing all irrelevant algae
     is.null(ncol(current_matrix[, which(colSums(current_matrix) != 0)]))){ 
    quadrat_level_portfolio[i, 2] <- NA # set portfolio effects to 0
  }else{ # otherwise
    current_matrix <- current_matrix[, which(colSums(current_matrix) != 0)] # remove irrelevant algae

@@

# sum(is.na(quadrat_level_portfolio$Portfolio_effect)) # 4 quadrats causing trouble

hist(quadrat_level_portfolio$Portfolio_effect) # some crazy outliers due to low richness

alpha_diversity_quad_sub <- alpha_diversity_quad[!alpha_diversity_quad$Location %in% problem_children, ]

alpha_diversity_quad_sub$Cover <- rowSums(data_sub[ , -which(names(data_sub) %in% meta_cols)])/100
alpha_diversity_quad_sub$Stratum <- 
  paste(alpha_diversity_quad_sub$Site, "-", alpha_diversity_quad_sub$Habitat, "-", alpha_diversity_quad_sub$Transect)
lm_synchrony_quadrat <- codyn::synchrony(df = long_data_sub,
                                   time.var = "Year",
                                   species.var = "Taxa",
                                   abundance.var = "Percent_cover",
                                   metric = "Loreau",
                                   replicate.var = "Location") # Loreau and Mazancourt synchrony
colnames(lm_synchrony_quadrat) = c("Location", "Loreau_synchrony") # rename cols for merge later

gross_synchrony_quadrat <- codyn::synchrony(df = long_data_sub,
                                   time.var = "Year",
                                   species.var = "Taxa",
                                   abundance.var = "Percent_cover",
                                   metric = "Gross",
                                   replicate.var = "Location") # lots of warnings in this one - examine further later
colnames(gross_synchrony_transect) = c("Location", "Gross_synchrony") # rename cols for merge later

diversity_stability_quadrat <- alpha_diversity_quad_sub %>%
  dplyr::group_by(Location, Site, Habitat, Site_habitat, Transect, Stratum) %>%
  dplyr::summarise(Richness_mean = mean(Richness), 
            Richness_se = sd(Richness)/sqrt(n()),
            Richness_cv = CV(Richness),
            Shannon_mean = mean(Shannon), 
            Shannon_se = sd(Shannon)/sqrt(n()),
            Shannon_cv = CV(Shannon),
            Shannon_stability = 1/CV(Shannon), # Stability as defined by codyn and many others (including Pete in 2022 paper)
            Cover_mean = mean(Cover),
            Cover_se = sd(Cover)/sqrt(n()),
            Cover_cv = CV(Cover),
            Cover_stability = 1/CV(Cover)) # as above

diversity_stability_quadrat <- 
  merge(diversity_stability_quadrat, lm_synchrony_quadrat, by = "Location") # add in loreau and de Mazancourt synchrony
diversity_stability_quadrat <- merge(diversity_stability_quadrat, gross_synchrony_quadrat, by = "Location")
diversity_stability_quadrat <- merge(diversity_stability_quadrat, quadrat_level_portfolio, by = "Location") # add in portfolio effects
pairs(~ Cover_mean + Cover_cv + Cover_stability + 
        Shannon_mean + Shannon_cv + Shannon_stability + 
        Loreau_synchrony + Gross_synchrony + Portfolio_effect, data = diversity_stability)

@@

```{r basic diversity stability plots quadrat level, fig.height=8, fig.width=10}
(shannon_cv_plot = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 8) + 
   scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
#    stat_smooth(aes(x = Shannon_mean, y = Shannon_cv),method = "lm", se = T, size = 2) + 
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

(shannon_cv_plot_2 = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

# LTER 2 obscures a lot of the trends

```{r portfolio effects and synchrony transect level,  fig.height = 8, fig.width = 10}
(loreau_stability_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Loreau_synchrony, y = Cover_cv, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

# (gross_stability_plot_quad = 
#   ggplot(data = diversity_stability_quadrat) +
#   geom_point(aes(x = Gross_synchrony, y = Cover_stability, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
#    scale_colour_manual(values = site_colours) +
#    scale_shape_manual(values = habitat_shapes) +
#   facet_wrap(~Site) +
#    
#   theme_bw() + 
#   theme(axis.title.x = element_text(size=18), 
#         axis.title.y = element_text(size=18), 
#         panel.background = element_blank(), 
#         panel.grid.major = element_blank(),  #remove major-grid labels
#         panel.grid.minor = element_blank(),  #remove minor-grid labels
#         plot.background = element_blank(),
#         legend.text = element_text(size = rel(1)), # increase legend text
#         legend.title = element_text(size = rel(1)),
#         legend.position = 'bottom', # move legend to bottom
#         legend.box = 'horizontal', # make legend horizontal
#         legend.box.background = element_rect(colour = "black"))  # add box around legend
#          )

(portfolio_stability_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Portfolio_effect, y = Cover_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

```{r diversity portfolio effects and synchrony quadrat level,  fig.height = 8, fig.width = 10}
(loreau_diversity_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Loreau_synchrony, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

(portfolio_diversity_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Portfolio_effect, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
